---
title: "WQD7004 Predicting public transport ridership in Kuala Lumpur based on various factors"
author: Mark Nicholas Akyapogu, Na Juinn Sheng, Nisyhaal Tamil A/L TamilChelvam, Tan Fong Yi, Xiao Rui Jie, Zhang Yanyubo
output: html_document
occurence: 2
Group: 
date: "2023-12-26"
---

# **Title: Predicting public transport ridership in Kuala Lumpur based on various factors **

## Introduction

According to a recent article by NST, Malaysia Transport Minister Anthony Loke has informed that cars made up the highest number of registered vehicles, with 17,244,978 cars in 2023. The total number of registered vehicles in Malaysia has reached more than 36.3 million, exceeding the country's population. The large number of cars on the road can be concerning as many vehicles may lead to heavy traffic congestion in the Klang Valley area. Besides, there is also a large amount of CO2 emissions due to emissions from such vehicles. Transport is Malaysia's second largest carbon dioxide driver after electricity and 
heat production (Solaymani, 2022). The lack of low-carbon technologies, the electrification of vehicles, the lack of EV charging ports widely implemented across Malaysia, and the lack of connectivity of rail networks and buses across the Klang Valley and other rural states are concerning. We aim to explore the main driving factors that could contribute to the fluctuating patterns of public transport ridership based on factors such as weather, Covid-19, fuel price, and public holidays and predict such ridership for each Prasarana rail and bus service. At the end of our project, we aim to contribute to infrastructure optimization - for instance, increasing intervals during the rainy season and using our data to support decision-making processes.

## Business Objective
1. To develop a time series forecasting model using Prasarana dataset obtained from data.gov.my 
2. To employ regression modelling by considering various factors that may influence the ridership of Prasarana services with an accuracy of 70%.
3. To explore and analyze the data using various EDA techniques to understand patterns and hidden insights in the dataset

## Understanding about our dataset
Our dataset is obtained from various sources, which include Wikipedia, data.gov.my, weather (visualcrossing.com), and publicholidays.com.my.
The ridership and fuel dataset obtained from data.gov.my has the most updated data as of the date we extracted the data from the website (2019 - November 2023), while the other datasets are obtained from 2019-2023.
The purpose of the dataset is to investigate the various contributing factors that may contribute to a spike or decrease in Prasarana ridership. We will explore the data's dimensions, content, structure and summary in the section below.


## References
https://www.nst.com.my/news/nation/2023/12/987062/363-million-vehicles-malaysia#
Solaymani, S. (2022). CO2 emissions and the transport sector in Malaysia. Frontiers in Environmental Science, 9, 774164. https://doi.org/10.3389/fenvs.2021.774164

## Team contribution
1. Na Juinn Sheng (Leader) - Introduction, Business Objective, Web Scraping, Obtaining Data, Merging of Data, Data Pre-processing, EDA Q19 - Q24, 3 Machine Learning Models (SARIMA, XGBOOST using Caret and MLR, K-Means Clustering), Video Editing + Presentation
2. Tan Fong Yi - Most of the EDA (EDA Q1 - Q18) + Presentation 
3. Nisyhaal Tamil - Machine Learning Model for every Prasarana services (Multiple Linear Regression, ARIMA) + Presentation
4. Zhang Yanyubo - EDA Q25 - Q29 + Presentation
5. Mark Nicholas Akyapogu - Machine Learning Model (Random Forest using Caret) + Presentation
6. Xiao Rui Jie - 1 box plot graph + Presentation on Introduction and Business Objective.

## Web scraping
```{r}

library(rvest)
library(tidyverse)
library(dplyr)

#This code section scrapes the Wikipedia table to obtain the MCO dates

url <- "https://en.wikipedia.org/wiki/Malaysian_movement_control_order"
html <- read_html(url)
html %>%
  html_elements(".wikitable") %>%
  html_table() -> wikitables
#This code select the relevant table
df <- wikitables[[1]]
df <- as.data.frame(wikitables[[1]])
df$Date <- as.character(df$Date)

#This code extracts the start date and end date
df <- df %>%
  mutate(Start_Date = str_extract(Date, "\\d{1,2} [a-zA-Z]+ \\d{4}"),
         End_Date = str_extract(Date, "(?<=– )\\d{1,2} [a-zA-Z]+ \\d{4}"))

df$Start_Date <- as.Date(df$Start_Date, format="%d %B %Y")
df$End_Date <- as.Date(df$End_Date, format="%d %B %Y")

df_Start_Date <- data.frame(Start_Date = df$Start_Date)

#We drop the duplicates and NaN values
df_Start_Date <- df_Start_Date %>%
  distinct(Start_Date) %>%
  filter(!is.na(Start_Date))

df_End_Date <- data.frame(End_Date = df$End_Date) %>%
  distinct(End_Date) %>%
  filter(!is.na(End_Date))

#Renaming the columns
colnames(df_Start_Date) <- colnames(df_End_Date)

#Combining df_Start_Date and df_End_Date
df_appended <- rbind(df_Start_Date, df_End_Date)
colnames(df_appended) <- "Covid_19_dates"
df_appended$Covid_19_dates <- as.Date(df_appended$Covid_19_dates, format="%d %B %Y")
#We arrange the dataframe from earliest date to latest
df_appended <- df_appended[order(df_appended$Covid_19_dates, decreasing = FALSE), ]
df_appended <- as.data.frame(df_appended)

colnames(df_appended) <- "Covid_19_dates"
#All these dates are not Kuala Lumpur and Selangor MCO dates. Hence we drop them
dates_to_drop <- as.Date(c("2021-07-05", "2021-07-07", "2021-07-10", "2021-07-14", "2021-08-26", "2021-09-04", "2021-09-24", "2021-08-04", "2021-09-04", "2021-09-17", "2021-09-24", "2021-10-08", "	
2021-06-15", "2021-08-26", "2021-09-24", "2021-10-08", "2021-10-25", "2021-11-08", "2022-01-03", "2021-06-28", "2021-12-31", "2021-06-15"))

df_appended <- df_appended %>%
  distinct(Covid_19_dates) %>%
  filter(!Covid_19_dates %in% dates_to_drop)

View(df_appended)
write.csv(df_appended, "Covid-19_dates.csv", row.names = FALSE)
```

Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot.


```{r}

#scaping 2021 holiday list
library(rvest)
library(dplyr)

# URL_2021
url_2021 <- "https://publicholidays.com.my/2021-dates/"

webpage_2021 <- read_html(url_2021)

#Selecting the relevant table from the website
table_data_2021 <- webpage_2021 %>%
  html_nodes("table.publicholidays") %>%
  html_table(fill = TRUE)

#Converting table to dataframe
table_data_2021 <- as.data.frame(table_data_2021[[1]])

#Extracting the holiday dates
table_data_2021 <- table_data_2021 %>%
  filter(grepl("^\\d+\\s\\w+$", Date))

table_data_2021 <- table_data_2021 %>%
  mutate(formatted_date = format(as.Date(Date, format = "%d %b"), "2021-%m-%d"))
View(table_data_2021)

holiday_2021 <- as.data.frame(table_data_2021$formatted_date)
View(holiday_2021)
colnames(holiday_2021) <- "holiday_2021_dates"
write.csv(holiday_2021, "Public_Holiday_2021.csv", row.names = FALSE)

```

```{r}
#scaping 2022 holiday
library(rvest)
library(dplyr)

# URL_2022
url_2022 <- "https://publicholidays.com.my/2022-dates/"

webpage_2022 <- read_html(url_2022)

table_data_2022 <- webpage_2022 %>%
  html_nodes("table.publicholidays") %>%
  html_table(fill = TRUE)

table_data_2022 <- as.data.frame(table_data_2022[[1]])
table_data_2022 <- table_data_2022 %>%
  filter(grepl("^\\d+\\s\\w+$", Date))
table_data_2022 <- table_data_2022 %>%
  mutate(formatted_date = format(as.Date(Date, format = "%d %b"), "2022-%m-%d"))
View(table_data_2022)

holiday_2022 <- as.data.frame(table_data_2022$formatted_date)
View(holiday_2022)
colnames(holiday_2022) <- "holiday_2022_dates"
write.csv(holiday_2022, "Public_Holiday_2022.csv", row.names = FALSE)
```

```{r}
#scaping 2023 holiday
library(rvest)
library(dplyr)

# URL_2023
url_2023 <- "https://publicholidays.com.my/2023-dates/"

webpage_2023 <- read_html(url_2023)

table_data_2023 <- webpage_2023 %>%
  html_nodes("table.publicholidays") %>%
  html_table(fill = TRUE)

table_data_2023 <- as.data.frame(table_data_2023[[1]])
table_data_2023 <- table_data_2023 %>%
  filter(grepl("^\\d+\\s\\w+$", Date))
table_data_2023 <- table_data_2023 %>%
  mutate(formatted_date = format(as.Date(Date, format = "%d %b"), "2023-%m-%d"))
View(table_data_2023)

holiday_2023 <- as.data.frame(table_data_2023$formatted_date)
View(holiday_2023)
colnames(holiday_2023) <- "holiday_2023_dates"
write.csv(holiday_2023, "Public_Holiday_2023.csv", row.names = FALSE)
```

```{r}
#scaping 2020 holiday
library(rvest)
library(dplyr)

# URL_2020
url_2020 <- "https://publicholidays.com.my/2020-dates/"

webpage_2020 <- read_html(url_2020)

table_data_2020 <- webpage_2020 %>%
  html_nodes("table.publicholidays") %>%
  html_table(fill = TRUE)

table_data_2020 <- as.data.frame(table_data_2020[[1]])
table_data_2020 <- table_data_2020 %>%
  filter(grepl("^\\d+\\s\\w+$", Date))
table_data_2020 <- table_data_2020 %>%
  mutate(formatted_date = format(as.Date(Date, format = "%d %b"), "2020-%m-%d"))
View(table_data_2020)

holiday_2020 <- as.data.frame(table_data_2020$formatted_date)
View(holiday_2020)
colnames(holiday_2020) <- "holiday_2020_dates"
write.csv(holiday_2020, "Public_Holiday_2020.csv", row.names = FALSE)
```

```{r}
#scaping 2019 holiday
library(rvest)
library(dplyr)

# URL_2019
url_2019 <- "https://publicholidays.com.my/2019-dates/"

webpage_2019 <- read_html(url_2019)

table_data_2019 <- webpage_2019 %>%
  html_nodes("table.publicholidays") %>%
  html_table(fill = TRUE)

table_data_2019 <- as.data.frame(table_data_2019[[1]])
table_data_2019 <- table_data_2019 %>%
  filter(grepl("^\\d+\\s\\w+$", Date))
table_data_2019 <- table_data_2019 %>%
  mutate(formatted_date = format(as.Date(Date, format = "%d %b"), "2019-%m-%d"))
View(table_data_2019)

holiday_2019 <- as.data.frame(table_data_2019$formatted_date)
View(holiday_2019)
colnames(holiday_2019) <- "holiday_2019_dates"
write.csv(holiday_2019, "Public_Holiday_2019.csv", row.names = FALSE)
```


```{r}
#Importing the csv
library(tidyr)
library(zoo)
library(stringr)

ridership <- read.csv('ridership_headline.csv')
fuelprice <- read.csv('fuelprice.csv')

#Select the series type = 'level' in fuel price
fuelprice_filtered <- fuelprice[fuelprice$series_type == 'level', c("date","ron95", "ron97", "diesel")]
weather <- read.csv('weather.csv')

#Renaming the weather and holiday column to date
names(weather)[names(weather) == 'datetime'] <- 'date'
colnames(holiday_2019) <- 'date'
colnames(holiday_2020) <- 'date'
colnames(holiday_2021) <- 'date'
colnames(holiday_2022) <- 'date'
colnames(holiday_2023) <- 'date'
colnames(df_appended) <- 'date'

#Extracting relevant features for weather
weather2 <- weather[, c('date', 'preciptype', 'temp', 'uvindex')]

listofholiday <- list(
  holiday_2019 = holiday_2019,
  holiday_2020 = holiday_2020,
  holiday_2021 = holiday_2021,
  holiday_2022 = holiday_2022,
  holiday_2023 = holiday_2023
)

#Combining all the list of holiday dataframe
holiday_df <- do.call(bind_rows, listofholiday)

#If holiday, create a column called holiday and encode 1 when it is holiday session
holiday_df$holiday <- ifelse(!is.na(holiday_df$date), 1, 0)
View(holiday_df)

#If covid, create a column called covid and encode 1 when it is covid period
df_appended$covid <- ifelse(!is.na(df_appended$date), 1, 0)
df_appended <- df_appended[, c("date", "covid")]


listofdf <- list(
  ridership = ridership,
  fuelprice = fuelprice_filtered,
  weather = weather2,
  holiday = holiday_df,
  covid = df_appended
)

#Using Reduce function to merge the multiple dataframe
result_df1 <- Reduce(function(x, y) merge(x, y, by = "date", all.x = TRUE), listofdf)


result_df1$ron95 <- na.locf(result_df1$ron95)
result_df1$ron97 <- na.locf(result_df1$ron97)
result_df1$diesel <- na.locf(result_df1$diesel)

first_filled_index <- which(result_df1$covid == 1)[1]
last_non_missing_index <- tail(which(!is.na(result_df1$covid)), 1)

result_df1$covid <- ifelse(seq_along(result_df1$covid) >= first_filled_index & seq_along(result_df1$covid) <= last_non_missing_index, 1, result_df1$covid)

View(result_df1)

write.csv(result_df1, "result_df.csv", row.names = FALSE)
```

## Data preprocessing
```{r}
library(dplyr)
result_df1 <- read.csv("result_df.csv")
#Selecting only Kuala Lumpur line
result_df_1 <- result_df1 %>%
  select(-c("bus_rkn", "bus_rpn", "rail_ets", "rail_intercity", "rail_komuter", "rail_tebrau"))


write.csv(result_df_1, 'result_df_1.csv', row.names=FALSE)

cname <- names(result_df_1)
for (i in cname){
  result_df_1[!is.na(result_df_1[i]) & result_df_1[i] == "", i] <- NA
}

result_df_1$covid <- ifelse(is.na(result_df_1$covid), 0, result_df_1$covid)
result_df_1$holiday <- ifelse(is.na(result_df_1$holiday), 0, result_df_1$holiday)
result_df_1$preciptype <- ifelse(result_df_1$preciptype == 'rain', 1, ifelse(is.na(result_df_1$preciptype), 0, result_df_1$preciptype))


#Fill in missing values
result_df_1$bus_rkl[is.na(result_df_1$bus_rkl)] <- mean(result_df_1$bus_rkl, na.rm = TRUE)
result_df_1$rail_mrt_pjy[is.na(result_df_1$rail_mrt_pjy)] <- mean(result_df_1$rail_mrt_pjy, na.rm = TRUE)
result_df_1$preciptype <- ifelse(is.na(result_df_1$preciptype), 0, result_df_1$preciptype)

#Check for missing values
cname <- names(result_df_1)
for (i in cname){
  print(paste(i, sum(result_df_1[i] == "")))
  result_df_1[!is.na(result_df_1[i]) & result_df_1[i] == "", i] <- NA
}

#Check for dimension, structure and summary of the data
str(result_df_1) #structure
summary(result_df_1) #summary
dim(result_df_1) #dimension

View(result_df_1)
write.csv(result_df_1, "df_cleaned.csv", row.names=FALSE)
```

## Exploratory Data Analysis
```{r library, echo=TRUE, include=FALSE}
library(ggplot2)
library(tidyverse)
library(ragg)
#library(palmerpenguins)
library(Hmisc)
library(corrplot)
library(GGally)
theme_update(plot.title = element_text(hjust = 0.5))

data<-read.csv("df_cleaned.csv")
str(data)

```
## 1. Reading data from file

```{r}
data<-read.csv("df_cleaned.csv")
str(data)
```

## 2. Transform the data into appropriate type

```{r}
data$date <-as.Date(data$date) # from char to date
data$bus_rkl<-as.integer(data$bus_rkl) # from num to int
data$rail_mrt_pjy<-as.integer(data$rail_mrt_pjy) # from num to int
str(data)
```

## 3. Takes public transports ridership columns and gathers them into key-value pairs, duplicating all other columns

```{r gather}
data1<-data %>%
  select("date", "bus_rkl", "rail_lrt_ampang", "rail_mrt_kajang", "rail_lrt_kj", "rail_monorail", "rail_mrt_pjy") %>%
  gather("transport", "passenger", 2:7)
head(data1,10)
```
## 4. Plot individual scatterplot for public transports ridership

```{r echo=TRUE, fig.height=12, fig.width=25}
plotname<-c('bus_rkl'='Bus', 'rail_lrt_ampang'="LRT Ampang",    "rail_mrt_kajang"="MRT Kajang", "rail_lrt_kj"="LRT Kelana Jaya", 'rail_monorail'='Monorail', 'rail_mrt_pjy'="MRT Putrajaya")
p1<- ggplot(data1, aes(x = date, y = (passenger/1000))) + 
     geom_point(color = "steel blue", size=1.5) + 
     facet_wrap(~transport, labeller = as_labeller(plotname)) +
     ggtitle("Public Transport Services Ridership") + xlab("Date") +  ylab("Daily ridership ('000)")+
     theme(axis.text=element_text(size=20), #change font size of axis text
           axis.title=element_text(size=20), #change font size of axis titles
           plot.title=element_text(size=30), #change font size of plot title
           legend.text=element_text(size=20), #change font size of legend text
           legend.title=element_text(size=20),
           strip.text = element_text(size=20)) #change font size of legend title  
plot(p1)
```

## 5. Plot multiple scatterplot for public transports ridership

```{r echo=TRUE, fig.height=10, fig.width=25}
p2<-ggplot(data1, aes(x = date, y = (passenger/1000), shape=transport, color=transport )) + 
  geom_point() + 
  ggtitle("Public Transport Services Ridership") + xlab("Date") + ylab("Daily ridership ('000)") +
  scale_color_manual(name = "Public transport", 
                     labels = c("Bus", "LRT Ampang","LRT Kelana Jaya", "Monorail","MRT Kajang","MRT Putrajaya"),
                     values = c("darkred", "steelblue", "yellow", 'purple','black','green'))+
  scale_shape_manual(values=c(15, 16, 17, 18, 4, 7),
                     labels = c("Bus", "LRT Ampang","LRT Kelana Jaya", "Monorail","MRT Kajang","MRT Putrajaya"),
                     name = "Public transport")+
  theme(axis.text=element_text(size=20), #change font size of axis text
        axis.title=element_text(size=20), #change font size of axis titles
        plot.title=element_text(size=30), #change font size of plot title
        legend.text=element_text(size=20), #change font size of legend text
        legend.title=element_text(size=20))+
        guides(color = guide_legend(override.aes = list(size = 5)))
plot(p2)
```

## 6. Plot pie chart of Public transport services total ridership

```{r echo=TRUE, fig.height=6, fig.width=6}
data2<- data.frame("Transport" = c("Bus", "LRT Ampang", "MRT Kajang", "LRT Kelana Jaya", "Monorail", "MRT Putrajaya"),
"Ridership" = c(sum(data$bus_rkl),sum(data$rail_lrt_ampang),sum(data$rail_mrt_kajang), sum(data$rail_lrt_kj), sum(data$rail_monorail),sum(data$rail_mrt_pjy)))

p4<-ggplot(data2, aes(x = "", y = Ridership, fill = Transport)) +
  geom_bar(width = 1, stat = "identity", color = "white") +
  coord_polar("y", start = 0)+
  geom_text(aes(label = Ridership), position = position_stack(vjust=0.5), color = "white")+
  ggtitle("Public Transport Services Total Ridership")+
  scale_fill_manual(values = c("darkred", "steelblue", "yellow", 'purple','black','green')) +
  labs(x = NULL, y = NULL) +
   theme_classic() +
  theme(plot.title=element_text(size=15, hjust = 0.5), 
        legend.text=element_text(size=8), 
        legend.title=element_text(size=10),
        axis.line = element_blank(),
        axis.text = element_blank(),
        axis.ticks = element_blank())
plot(p4)
```

## 7.Plot individual scatterplot for public transports ridership during holiday and non holiday
```{r echo=TRUE, fig.height=12, fig.width=30}
data3<-data %>%
  gather("transport", "passenger", 2:7)
data3$holiday<-as.character(data3$holiday)

p5<- ggplot(data3, aes(x=date, y = (passenger/1000), group=holiday)) + 
  geom_point(aes(color=holiday), size=1) +
  theme_gray(base_size = 10) +
  facet_wrap(~transport, labeller = as_labeller(plotname), scales="free") +
  ggtitle("Public Transport Services Ridership During Holiday and Non Holiday") + xlab("Date") + ylab("Daily ridership ('000)") +
  scale_color_manual(values = c("blue", "orange"),
                     name = "Holiday", 
                     labels = c("Non holiday",'holiday'))+
  theme(axis.text=element_text(size=20), #change font size of axis text
        axis.title=element_text(size=20), #change font size of axis titles
        plot.title=element_text(size=30, hjust = 0.5), #change font size of plot title
        legend.text=element_text(size=20), #change font size of legend text
        legend.title=element_text(size=20),
        strip.text = element_text(size=20))+
        guides(color = guide_legend(override.aes = list(size = 5)))
plot(p5)  
```

## 8. Plot individual histogram to show the distribution of public transports  ridership during covid and no covid
```{r echo=TRUE, fig.height=12, fig.width=30}
data3$covid<-as.character(data3$covid)

p6<- ggplot(data3, aes(x=(passenger/1000), fill=covid)) + 
  geom_histogram(position="identity", alpha=0.5, color='black', binwidth=5) +
  theme_gray(base_size = 10) +
  facet_wrap(~transport, labeller = as_labeller(plotname), scales='free') +
  ggtitle("Public Transport Services Ridership During Covid and Non Covid") + ylab("Count") + xlab("Daily ridership ('000)") +
   scale_fill_manual(name = "Covid", labels = c("Non Covid",'Covid'), 
                     values=c("deepskyblue1", "coral"))+
   theme(axis.text=element_text(size=20), #change font size of axis text
        axis.title=element_text(size=20), #change font size of axis titles
        plot.title=element_text(size=30, hjust = 0.5), 
        legend.text=element_text(size=20), #change font size of legend text
        legend.title=element_text(size=20),
        strip.text = element_text(size=20))
plot(p6) 

```

## 9. Plot individual scatterplot for public transports ridership based on preciptype
```{r echo=TRUE, fig.height=12, fig.width=30}
data3$preciptype<-as.character(data3$preciptype)

p7<- ggplot(data3, aes(x=date, y = (passenger/1000), group=preciptype)) + 
  geom_point(aes(color=preciptype), size=1.5) +
  theme_gray(base_size = 10) +
  facet_wrap(~transport, labeller = as_labeller(plotname), scales="free") +
  ggtitle("Public Transport Services Ridership") + xlab("Date") + ylab("Daily ridership ('000)") +
  scale_color_manual(values = c("blue", "darksalmon"),
                     name = "Precitype", 
                     labels = c("No rain",'Rain'))+
  theme(axis.text=element_text(size=20), #change font size of axis text
        axis.title=element_text(size=20), #change font size of axis titles
        plot.title=element_text(size=30, hjust = 0.5), #change font size of plot title
        legend.text=element_text(size=20), #change font size of legend text
        legend.title=element_text(size=20),
        strip.text = element_text(size=20))+
        guides(color = guide_legend(override.aes = list(size = 5)))
plot(p7) 
```
## 10. Plot scatterplot for daily ridership and temperature
```{r echo=TRUE, fig.height=12, fig.width=30}
p8<- ggplot(data3, aes(x = temp, y = (passenger/1000))) + 
     geom_point(color = "steel blue", size=1.5) + 
     facet_wrap(~transport, labeller = as_labeller(plotname), scales='free') +
     ggtitle("Public Transport Services Ridership and Temperature") + xlab("Temperature") +  ylab("Daily ridership ('000)")+
     theme(axis.text=element_text(size=20), #change font size of axis text
           axis.title=element_text(size=20), #change font size of axis titles
           plot.title=element_text(size=30), #change font size of plot title
           legend.text=element_text(size=20), #change font size of legend text
           legend.title=element_text(size=20),
           strip.text = element_text(size=20)) #change font size of legend title
plot(p8)
```

## 11. Plot scatterplot for daily ridership and UV Index
```{r echo=TRUE, fig.height=12, fig.width=30}
p9<- ggplot(data3, aes(x = uvindex, y = (passenger/1000))) + 
     geom_point(color = "steel blue", size=1.5) + 
     facet_wrap(~transport, labeller = as_labeller(plotname), scales='free') +
     ggtitle("Public Transport Services Ridership and Temperature") + xlab("UV index") +  ylab("Daily ridership ('000)")+
     theme(axis.text=element_text(size=20), #change font size of axis text
           axis.title=element_text(size=20), #change font size of axis titles
           plot.title=element_text(size=30), #change font size of plot title
           legend.text=element_text(size=20), #change font size of legend text
           legend.title=element_text(size=20),
           strip.text = element_text(size=20)) #change font size of legend title
plot(p9)
```

## 13. Plot scatterplot for daily ridership and gas prices
```{r echo=TRUE, fig.height=12, fig.width=30}
data4<-data3 %>%
  gather("Gas", "Prices", 2:4)

p10<- ggplot(data4, aes(x=Prices, y = (passenger/1000), group=Gas)) + 
  geom_point(aes(color=Gas), size=1.5) +
  theme_gray(base_size = 10) +
  facet_wrap(~transport, labeller = as_labeller(plotname), scales="free") +
  ggtitle("Public Transport Services Ridership") + xlab("Gas") + ylab("Daily ridership ('000)") +
  scale_color_manual(values = c("blue", "darksalmon", "yellow"),
                     name = "Gas", labels = c("Diesel",'Ron 95','Ron 97'))+
  theme(axis.text=element_text(size=20), #change font size of axis text
        axis.title=element_text(size=20), #change font size of axis titles
        plot.title=element_text(size=30, hjust = 0.5), #change font size of plot title
        legend.text=element_text(size=20), #change font size of legend text
        legend.title=element_text(size=20),
        strip.text = element_text(size=20))+
  guides(color = guide_legend(override.aes = list(size = 5)))
plot(p10) 
```

## 14. Statistical summary of the data
```{r echo=TRUE}
summary(data)
```

## 15. Plot individual histogram and density for public transports ridership
```{r echo=TRUE, fig.height=12, fig.width=30}
p11<-data3 %>%
  ggplot(aes(x=passenger/1000)) +
  geom_histogram(aes(y=after_stat(density)),color="steel blue", fill="steel blue", alpha=0.6, position = 'identity') +
  geom_density(color="dark red")+
  facet_wrap(~transport, labeller = as_labeller(plotname), scales='free') +
  ggtitle("Public Transport Services Ridership") + ylab('Density') + xlab("Ridership ('000)")+
  theme(axis.text=element_text(size=20), #change font size of axis text
        axis.title=element_text(size=20), #change font size of axis titles
        plot.title=element_text(size=30),
        strip.text = element_text(size=20)) #change font size of plot title
plot(p11)

```

## 16. Correlation coefficient and p-value matrix

```{r echo=TRUE}
corr <- rcorr(as.matrix.data.frame(data[,2:15]))
corr
```

## 17. Creates a correlogram to display the correlation matrix
```{r echo=TRUE} 
corrplot(corr$r, type = "upper", order = "hclust", 
         tl.col = "black", tl.srt = 45)
```
## 18. What is the boxplot of the entire distribution?
```{r}
library(ggplot2)
library(dplyr)
library(tidyr)
df <- read.csv("df_cleaned.csv")
df <- df %>%
  select(-date)  # Remove the 'df,' prefix

column_names <- colnames(df)
df2 <- gather(df)
ggplot(df2, aes(x = key, y = value)) +
  geom_boxplot(fill = "#0c4c8a") +
  theme_minimal() +
  labs(x = "Column Name", y = "Value", title = "Boxplot for the Entire Data Frame") + theme(axis.text.x = element_text(angle = 45, hjust = 1))

```

## 19. Univariate analysis of distribution of LRT users at Kelana Jaya line
```{r}
library(ggplot2)
library(dplyr)
library(tidyr)
df <- read.csv("df_cleaned.csv")
df <- df %>%
  select(-date)
column_names <- colnames(df)
for (i in column_names){
  histogram <- ggplot(data=df, aes(x = rail_lrt_kj)) + geom_histogram()
  print(histogram)
}

```
## 20. How is ridership distributed between different fuel types for rail_lrt_kj?
```{r}
df3 <- gather(df, key = "Fuel_Type", value = "Fuel_Price", ron95, ron97, diesel)

ggplot(df3, aes(x = Fuel_Price, y = rail_lrt_kj, color = Fuel_Type)) +
  geom_point() +
  labs(x = "Fuel Price", y = "Rail LRT KJ Ridership", color = "Fuel Type") +
  ggtitle("Scatter Plot of Rail LRT KJ Ridership vs. Fuel Price")

```
## 21. Is there an increase in ridership when it is raining?
```{r}

library(ggplot2)
df4 <- df
df4$preciptype <- ifelse(df4$preciptype == 1, "Rain", "No Rain")
ggplot(df4, aes(x = preciptype, y = rail_lrt_kj)) +
  geom_bar(stat="identity") +
  labs(x = "Precipitation Type", y = "Ridership", title = "Ridership when it is raining") +
  theme_minimal()
```
## 22. What is the correlation between the variables for machine learning model?
```{r}

library(e1071)
library(corrplot)
library(xgboost)
library(cowplot)

corr <- cor(df)
corrplot(corr, method='number', type="upper", order="hclust", addCoef.col =  "black")
```

## 23. How is the ridership of LRT Kelana Jaya over time?
```{r}

library(ggplot2)
library(dplyr)
df <- read.csv("df_cleaned.csv")
cutoff_date_start <- as.Date("2020-03-01")
cutoff_date_end <- as.Date("2020-12-31")
df %>%
  select(date, rail_lrt_kj, rail_mrt_kajang, rail_lrt_ampang, rail_monorail, rail_mrt_pjy) %>%
  gather(key = "rail_type", value = "ridership", -date) %>%
  mutate(date = as.Date(date)) %>%
  ggplot(aes(x = date, y = ridership, color = rail_type)) +
  geom_line() +
  geom_vline(xintercept = as.numeric(cutoff_date_start), linetype = "dashed", color = "red") +
  geom_vline(xintercept = as.numeric(cutoff_date_end), linetype = "dashed", color = "red") +
  labs(x = "Date", y = "Ridership", color = "Rail Type") +
  ggtitle("Line Graph of Rail LRT/Monorail/MRT Over Time")
```
## 24.What is the ridership comparison between holiday and non-holiday?
```{r}

df <- read.csv("df_cleaned.csv")
df4 <- df %>%
  mutate(date = as.Date(date)) %>%
  filter(covid == 1)
df4 <- df %>%
  mutate(holiday = ifelse(holiday == 1, "Holiday", "Non-Holiday"))

ridership_summary <- df %>%
  group_by(holiday) %>%
  summarise(Average_Ridership = mean(rail_lrt_kj, na.rm = TRUE))

ggplot(ridership_summary, aes(x = holiday, y = Average_Ridership)) +
  geom_bar(stat = "identity", position = "dodge") +
  labs(title = "Ridership Comparison between Holidays and Non-Holidays",
       x = "Day Type", y = "Average Ridership") +
  theme_minimal()

```

## 25.Rail Monorail Passengers Over Time

```{r}
df <- read_csv('df_cleaned.csv')
df$date <- as.Date(df$date)
library(ggplot2)
ggplot(df, aes(x = date, y = rail_monorail)) +
  geom_line() +
  labs(title = "Rail Monorail Passengers Over Time", x = "Date", y = "Passengers")
```

##  26. Average UV Index by Holiday

```{r uv-index-holiday-bar}
ggplot(df, aes(x = factor(holiday), y = uvindex)) +
  geom_bar(stat = "summary", fun = "mean") +
  labs(title = "Average UV Index by Holiday", x = "Holiday", y = "Average UV Index")
```
## 27. Boxplot of Temperatures
```{r temperature-boxplot}
ggplot(df, aes(y = temp)) +
  geom_boxplot() +
  labs(title = "Boxplot of Temperatures", y = "Temperature")
```

## 28. Diesel Prices vs. Bus Ridership

```{r diesel-bus-scatter}
ggplot(df, aes(x = diesel, y = bus_rkl)) +
  geom_point() +
  labs(title = "Diesel Prices vs. Bus Ridership", x = "Diesel Price", y = "Bus Ridership")
```

##  29. Histogram of Rail MRT Kajang Passengers
```{r}
ggplot(df, aes(x = rail_mrt_kajang)) +
  geom_histogram(bins = 30, fill = "blue", color = "white") +
  labs(title = "Histogram of Rail MRT Kajang Passengers", x = "Passengers")
```

## Data Preprocessing for SARIMA
```{r}
df_keras <- read.csv("df_cleaned.csv")
df_keras_preprocessing <- data.matrix(df_keras)
mean <- apply(df_keras_preprocessing, 2, mean)
std <- apply(df_keras_preprocessing, 2, sd)
df_keras_preprocessing <- scale(df_keras_preprocessing, center=mean, scale=std)

#Normalize function
normalize <- function(x){
  return ((x - min(x) / max(x) - min(x)))
}
max <- apply(df_keras_preprocessing, 2, max)
min <- apply(df_keras_preprocessing, 2, min)

#Visualize the normalize data
df_keras_preprocessing <- apply(df_keras_preprocessing, 2, normalize)
plot(df_keras_preprocessing[1: 1000, 4])
```


## SARIMA forecast with categorising predictions into low, medium and high ridership
```{r}
library(forecast)
library(dplyr)
df_sarima <- read.csv("df_cleaned.csv")
df_sarima %>%
  select(date, rail_lrt_kj)

set.seed(123)


#Defining a time series object
daily_data <- ts(df_sarima$rail_lrt_kj, frequency = 365)  

sarima_model <- auto.arima(daily_data, seasonal = (period=365)) #select the best p,d, q values with lowest AIC

summary(sarima_model)

#Forecast for the next 30 days
forecast_values <- forecast(sarima_model, h = 30)  

#Categorise the predictions into low, medium, high
categories <- ifelse(forecast_values$mean < 100000, 'Low ridership',
                     ifelse(forecast_values$mean >= 100000 & forecast_values$mean < 300000, 'Medium ridership',
                            ifelse(forecast_values$mean >= 300000, 'High ridership', NA)))

#Display forecasted results in a dataframe
forecast_data <- data.frame(Date = seq(as.Date(Sys.Date()) + 1, by = "days", length.out = length(forecast_values$mean)),
                            Forecast = forecast_values$mean,
                            Category = categories)
View(forecast_data)

#Plot the forecast values

plot(forecast_values, main = "SARIMA Forecast")
lines(daily_data, col = "black", lty = 1)
```

## Multiple Linear Regression Preprocessing
```{r}
library(caret)
library(ggplot2)
# Import dataset
data <- read.csv("df_cleaned.csv")

# Drop the 'date' column
data <- data[, -1]

# Convert 'preciptype', 'holiday', and 'covid' to factors
data$preciptype <- as.factor(data$preciptype)
data$holiday <- as.factor(data$holiday)
data$covid <- as.factor(data$covid)

# Normalize the predictors
data_normalized <- data

# Exclude non-numeric columns from normalization
numeric_cols <- sapply(data_normalized, is.numeric)
data_normalized[, numeric_cols] <- scale(data_normalized[, numeric_cols])

set.seed(123)  # For reproducibility
train_index <- createDataPartition(data_normalized$bus_rkl, p = 0.9, list = FALSE)
train_data <- data_normalized[train_index, ]
test_data <- data_normalized[-train_index, ]
```

## Multiple Linear Regression on Bus RKL
```{r}
# Linear regression model training
model_bus_rkl <- lm(bus_rkl ~ ., data = train_data)

# Make predictions
predictions_bus_rkl <- predict(model_bus_rkl, newdata = test_data)

# Model evaluation
bus_rkl_metrics <- caret::defaultSummary(data.frame(obs = test_data$bus_rkl, pred = predictions_bus_rkl))
print("Normalized Data Metrics (Bus RKL):")
print(bus_rkl_metrics)
```

## Multiple Linear Regression on LRT Ampang
```{r}
# Linear regression model training
model_rail_lrt_ampang <- lm(rail_lrt_ampang ~ ., data = train_data)

# Make predictions
predictions_rail_lrt_ampang <- predict(model_rail_lrt_ampang, newdata = test_data)

# Model evaluation
rail_lrt_ampang_metrics <- caret::defaultSummary(data.frame(obs = test_data$rail_lrt_ampang, pred = predictions_rail_lrt_ampang))
print("Normalized Data Metrics (LRT Ampang):")
print(rail_lrt_ampang_metrics)
```

## Multiple Linear Regression on MRT Kajang
```{r}
# Linear regression model training
model_rail_mrt_kajang <- lm(rail_mrt_kajang ~ ., data = train_data)

# Make predictions
predictions_rail_mrt_kajang <- predict(model_rail_mrt_kajang, newdata = test_data)

# Model evaluation
rail_mrt_kajang_metrics <- caret::defaultSummary(data.frame(obs = test_data$rail_mrt_kajang, pred = predictions_rail_mrt_kajang))
print("Normalized Data Metrics (MRT Kajang):")
print(rail_mrt_kajang_metrics)
```

## Multiple Linear Regression on LRT Kelana Jaya
```{r}
# Linear regression model training
model_rail_lrt_kj <- lm(rail_lrt_kj ~ ., data = train_data)

# Make predictions
predictions_rail_lrt_kj <- predict(model_rail_lrt_kj, newdata = test_data)

# Model evaluation
rail_lrt_kj_metrics <- caret::defaultSummary(data.frame(obs = test_data$rail_lrt_kj, pred = predictions_rail_lrt_kj))
print("Normalized Data Metrics (LRT Kelana Jaya):")
print(rail_lrt_kj_metrics)
```

## Multiple Linear Regression on Monorail
```{r}
# Linear regression model training
model_rail_monorail <- lm(rail_monorail ~ ., data = train_data)

# Make predictions
predictions_rail_monorail <- predict(model_rail_monorail, newdata = test_data)

# Model evaluation
rail_monorail_metrics <- caret::defaultSummary(data.frame(obs = test_data$rail_monorail, pred = predictions_rail_monorail))
print("Normalized Data Metrics (Monorail):")
print(rail_monorail_metrics)
```

## Multiple Linear Regression on MRT Putrajaya
```{r}
# Linear regression model training
model_rail_mrt_pjy <- lm(rail_mrt_pjy ~ ., data = train_data)

# Make predictions
predictions_rail_mrt_pjy <- predict(model_rail_mrt_pjy, newdata = test_data)

# Model evaluation
rail_mrt_pjy_metrics <- caret::defaultSummary(data.frame(obs = test_data$rail_mrt_pjy, pred = predictions_rail_mrt_pjy))
print("Normalized Data Metrics (MRT Putrajaya):")
print(rail_mrt_pjy_metrics)
```

## Code below is the ARIMA modeling for the dataset
```{r}
# Load libraries
library(tidyverse)
library(forecast)
```

## Load & Explore Data
```{r}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(forecast)
```

## Load & Explore Data for ARIMA
```{r}
data <- read.csv("df_cleaned.csv")
str(data)
```

## Data preprocessing for ARIMA
```{r}
library(dplyr)
# Convert 'date' column to a Date object
data$date <- as.Date(data$date)

# Convert 'preciptype', 'holiday', and 'covid' to factors
data$preciptype <- as.factor(data$preciptype)
data$holiday <- as.factor(data$holiday)
data$covid <- as.factor(data$covid)

# Normalization function
normalize_series <- function(series) {
  scale(series)
}

# Normalize the entire dataset
normalized_data <- data %>%
  mutate(across(where(is.numeric), normalize_series))
```

## Data Splitting for ARIMA
```{r}
# Split the data into training and testing sets
train_ratio <- 0.8  # 80% for training, 20% for testing
train_size <- floor(train_ratio * nrow(normalized_data))
train_data <- normalized_data[1:train_size, ]
test_data <- normalized_data[(train_size + 1):nrow(normalized_data), ]
```

## ARIMA Model Training and Evaluation for ARIMA
```{r}
library(forecast)
arima_model <- function(train, test, target_variable) {
  cat("Training ARIMA model for", target_variable, "\n")
  
  # Extract the target variable
  y_train <- train[[target_variable]]
  y_test <- test[[target_variable]]
  
  # Fit ARIMA model
  arima_fit <- auto.arima(y_train)
  
  # Forecast
  forecast_values <- forecast(arima_fit, h = length(y_test))
  
  # Model Evaluation
  accuracy_metrics <- accuracy(forecast_values, y_test)
  
  # Print evaluation metrics
  print(accuracy_metrics)
  
  # Plot actual vs. predicted
  plot(forecast_values, main = paste("ARIMA Forecast for", target_variable))
  lines(y_test, col = "red")
  
  # Save the model
  save(arima_fit, file = paste("arima_model_", target_variable, ".RData", sep = ""))
}

# Apply the function for each target variable
target_variables <- c("bus_rkl", "rail_lrt_ampang", "rail_mrt_kajang", "rail_lrt_kj", "rail_monorail", "rail_mrt_pjy")

for (target_var in target_variables) {
  arima_model(train_data, test_data, target_var)
}
```

## XgBoost on RapidKL Kelana Jaya line with caret package
```{r}
#loading libraries
library(xgboost)
library(Metrics)
library(caret)
library(dplyr)



#Removing data column
df <- read.csv("df_cleaned.csv")
df <- df %>%
  select(-date)


#Train-test-split the data based on 70% and 30%
?createDataPartition
set.seed(123)
trainIndex <- createDataPartition(df$rail_lrt_kj, p = 0.7, list=FALSE)
X <- df
y <- df %>%
  select(rail_lrt_kj)

X_train <- X[trainIndex, ]
X_test <- X[-trainIndex, ]
y_train <- y[trainIndex, ]
y_test <- y[-trainIndex, ]


#Step 1: Training Search Grid
grid_xg <- expand.grid(
  nrounds=100,
  max_depth=3,
  eta = 0.3,
  gamma=0,
  colsample_bytree=1,
  min_child_weight=1,
  subsample=1
)

#Step 2: Train Control

train_control <- caret::trainControl(
  method="none",
  verboseIter = FALSE,
  allowParallel = TRUE
)

#Step 3: Training

xgb <- caret::train(x = X_train, y = y_train, trControl = train_control, method = "xgbTree", tuneGrid=grid_xg)

#Step 4: Prediction

pred <- predict(xgb, X_test)
best_params <- xgb$bestTune


#Step 7: Evaluation of model

rsquared_xgboost <- cor(pred, y_test)^2
print(rsquared_xgboost)
xgboost_mae <- mae(pred, y_test)
xgboost_rmse <- rmse(pred, y_test)
cat("MAE of Xgboost is : ", xgboost_mae, "\n")
cat("RMSE of Xgboost is: ", xgboost_rmse, "\n")

```

## XgBoost on RapidKL Kelana Jaya line with mlr package
```{r}
#loading libraries
library(xgboost)
library(Metrics)
library(caret)
library(dplyr)
library(mlr)

#Removing data column
df <- read.csv("df_cleaned.csv")
df <- df %>%
  select(-date)

#Train-test-split the data based on 70% and 30%
?createDataPartition
set.seed(123)
trainIndex <- createDataPartition(df$rail_lrt_kj, p = 0.7, list=FALSE)
X <- df
y <- df %>%
  select(rail_lrt_kj)

X_train <- X[trainIndex, ]
X_test <- X[-trainIndex, ]
y_train <- y[trainIndex, ]
y_test <- y[-trainIndex, ]

?preProcess

#Step 1: Preprocessing and normalization
preprocess_params_xg <- preProcess(X_train, method = c("center", "scale"))
X_train_preprocessed <- predict(preprocess_params_xg, X_train)
X_test_preprocessed <- predict(preprocess_params_xg, X_test)

#Step 2: Specifying the parameters (Refer to documentation ?xgboost) - similar concept to param_grid for Python
param_set_xgboost <- makeParamSet(
  makeNumericParam("eta", lower = 0.01, upper = 0.3),
  makeIntegerParam("nrounds", lower = 50, upper = 120),
  makeIntegerParam("max_depth", lower = 3, upper = 9),
  makeIntegerParam("min_child_weight", lower = 1, upper = 5),
  makeNumericParam("subsample", lower = 0.8, upper = 1),
  makeNumericParam("colsample_bytree", lower = 0.8, upper = 1),
  makeNumericParam("lambda", lower = 0, upper = 5),
  makeNumericParam("alpha", lower = 0, upper = 1)
)


#Step 3: Creating the model
xgb <- makeLearner("regr.xgboost", predict.type="response")

#Step 4: Create a regression task
task <- makeRegrTask(data = X_train_preprocessed, target = "rail_lrt_kj")

#Step 5: Perform hyperparameter tuning, performing RandomisedSearch
res_xg <- tuneParams(
  xgb, #model 
  task = task, #regression task
  resampling = makeResampleDesc("CV", iters=3), #cross-validation = 3
  par.set = param_set_xgboost, #parameters defined as above
  control = makeTuneControlRandom(maxit=50) #randomly sample combinations of parameter and evaluate their performance for up to 50 iterations
)

#Step 6: Finding the best model
best_model_xg <- res_xg$learner 
final_model_xg <- train(best_model_xg, task)
y_pred_xg <- predict(final_model_xg, newdata = X_test_preprocessed)
inverse_transformed_predictions <- exp(y_pred_xg$data$truth) #because the data is normalized hence we need to inverse transform the prediction

#Step 7: Evaluation of model
rsquared_xgboost <- cor(inverse_transformed_predictions, y_test)^2
print(rsquared_xgboost)
plot(inverse_transformed_predictions, y_pred_xg$data$truth)
xgboost_mae <- mae(inverse_transformed_predictions, y_pred_xg$data$truth)
xgboost_rmse <- rmse(inverse_transformed_predictions, y_pred_xg$data$truth)
cat("MAE of Xgboost is : ", xgboost_mae, "\n")
cat("RMSE of Xgboost is: ", xgboost_rmse, "\n")
```

## K-means clustering on Rapid KL LRT Kelana Jaya Line
```{r}
library(dplyr)
library(caret)
library(mlr)
library(fpc)

#Defining X and y
set.seed(43)
X <- df %>%
  select(-rail_lrt_kj)
y <- df %>%
  select(rail_lrt_kj)

#Step 1: preprocessing
preprocess_params_k_cluster <- preProcess(X, method = c("center", "scale"))
X_preprocessed <- predict(preprocess_params_k_cluster, X)

#Step 2: Doing WSS score
wss <- NULL
for (i in 1:10){
  kmeans_model = kmeans(X_preprocessed, centers=i)
  wss[i] <- sum(kmeans_model$withinss)
}

plot(1:10, wss, type='o', xlab = "Number of clusters (k)", ylab = "Within the sum of Squares WSS", main = "Elbow Method to Determine Optimal k" )

k = 5 #defining the K

#Step 3: Plotting the k_means cluster
kmeans_model <- kmeans(X_preprocessed, centers=k)
plotcluster(X_preprocessed, kmeans_model$cluster, pointsbyclvecd = FALSE)

```

## Below code is for random forest using caret

```{r}
library(caret)
library(randomForest)
library(ggplot2)
library(readr)
library(dplyr)

data <- read.csv("df_cleaned.csv")

# Drop the 'date' column
data <- data[, -1]

# Convert 'preciptype', 'holiday', and 'covid' to factors
data$preciptype <- as.factor(data$preciptype)
data$holiday <- as.factor(data$holiday)
data$covid <- as.factor(data$covid)

# Normalize the predictors
data_normalized <- data

# Exclude non-numeric columns from normalization
numeric_cols <- sapply(data_normalized, is.numeric)
data_normalized[, numeric_cols] <- scale(data_normalized[, numeric_cols])

set.seed(123)  # For reproducibility
train_index <- createDataPartition(data_normalized$rail_lrt_kj, p = 0.9, list = FALSE)
train_data <- data_normalized[train_index, ]
test_data <- data_normalized[-train_index, ]
head(train_data)

rf <- randomForest(rail_lrt_kj ~ .,  # Formula specifying predictors
                   data = train_data,
                   ntree = 500,  # Number of trees (adjust as needed)
                   )

# Predictions
predictions <- predict(rf, test_data)

# Model Performance
lrt_kj_metrics <- caret::defaultSummary(data.frame(obs = test_data$rail_lrt_kj, pred = predictions))
print("Normalized Data Metrics (LRT Kelana Jaya Line):")
print(lrt_kj_metrics)


# Compare predictions and actual values
comparison <- data.frame(Actual = test_data$rail_lrt_kj, Predicted = predictions)

# Display the comparison
print("Comparison of Actual vs Predicted Values:")
print(comparison)
```

## Conclusion
Multiple Linear Regression (0.99) was able to demonstrate a high level of accuracy with low MAE and RMSE compared to Random Forest (0.98) and XgBoost (0.85).
